Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-01 papers/wdmerger_I/paper.tex
Added a reason for why we need a grid-based code to the introduction.

Around the same time, however, \cite{guillochon:2010} and \cite{dan:2011} came to significantly different conclusions. They pointed out that the previously mentioned simulations generally all shared a significant drawback, which was that their initial conditions were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006}, and \cite{motl:2007} (the first three-dimensional grid-based simulations of white dwarf mergers) pioneered the study of looking at the long-term dynamical evolution of merger scenarios after constructing exact equilibrium initial conditions. In contrast to the other studies, they found that paying close attention to the accuracy of the initial conditions of the simulation has important effects. In particular, earlier work placed the stars too close together and ignored the effects of tidal forces on changing the shape of the secondary, leading to the merger happening artificially too quickly. When the initial conditions are constructed in exact equilibrium, the system can be stable for tens of orbital periods, substantially changing the character of the mass transfer phase. However, one limitation of this series of studies is that the authors used a polytropic equation of state and thus could not consider nuclear reactions. \cite{guillochon:2010} and \cite{dan:2011} improved on this using a realistic equation of state, a nuclear reaction network, and a similar approach to the equilibrium initial conditions, and found substantial agreement with the idea that mass transfer occurs in a stable manner over tens of orbital periods. They also found that, assuming the material accreted onto the surface of the primary was primarily helium, explosive surface detonations would occur as a result of accretion stream instabilities during the mass transfer phase prior to the full merger. This could trigger a double-detonation explosion and thus perhaps a SN Ia.

@@

>>>>

Around the same time, however, \cite{guillochon:2010} and \cite{dan:2011} came to significantly different conclusions. They pointed out that the previously mentioned simulations generally all shared a significant drawback, which was that their initial conditions were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006}, and \cite{motl:2007} (the first three-dimensional grid-based simulations of mass transfer in binary white dwarf systems) pioneered the study of looking at the long-term dynamical evolution of binary white dwarf systems after constructing exact equilibrium initial conditions. In contrast to the other studies, they found that paying close attention to the accuracy of the initial conditions of the simulation has important effects. In particular, earlier work placed the stars too close together and ignored the effects of tidal forces on changing the shape of the secondary, leading to the merger happening artificially too quickly. When the initial conditions are constructed in exact equilibrium, the system can be stable for tens of orbital periods, substantially changing the character of the mass transfer phase. However, one limitation of this series of studies is that the authors used a polytropic equation of state and thus could not consider nuclear reactions. \cite{guillochon:2010} and \cite{dan:2011} improved on this using a realistic equation of state, a nuclear reaction network, and a similar approach to the equilibrium initial conditions, and found substantial agreement with the idea that mass transfer occurs in a stable manner over tens of orbital periods. They also found that, assuming the material accreted onto the surface of the primary was primarily helium, explosive surface detonations would occur as a result of accretion stream instabilities during the mass transfer phase prior to the full merger. This could trigger a double-detonation explosion and thus perhaps a SN Ia.

@@

Given the above, why is another approach using a different simulation code warranted? First and foremost, reproducibility of the results across simulation codes and algorithms is important for gauging confidence in this result. With only a couple of exceptions, most of the existing results that study the viability of the DD progenitor as an explanation for Type Ia supernovae (that is, including a temperature-dependent equation of state and nuclear reactions) have used the SPH method. SPH codes have a number of features which do aid them in the study of these systems, such as excellent conservation of angular momentum. However, the question of whether a prompt detonation in a merger actually happens depends in detail on the nature of the gas at the interface between the two stars, which is at much lower density than the rest of the stellar material. The SPH codes use uniform mass particles, so their effective resolution is \textit{lowest} at the stellar surface. In contrast, a grid-based code with adaptive mesh refinement can zoom in on the regions where hotspots will develop, while also maintaining high enough resolution in the high-density regions to adequately capture the large-scale mass transfer dynamics. However, there are also outstanding questions of convergence in SPH (e.g. \cite{zhu_SPH:2014}) and whether the method correctly captures fluid instabilities. This is an important question for white dwarf mergers because of the likely importance small-scale instabilities will have on the evolution of the low-density gas at the primary's surface. The pioneering work of \cite{agertz:2007} compared grid and SPH codes and found some important differences. Most relevant for this discussion is that the SPH codes could not adequately handle mixing from the Kelvin-Helmholtz instability in the test they propose. \cite{price:2008} attributed this to the fact that the standard SPH evolution equations do not have a mechanism for capturing discontinuities in internal energy, and showed that the addition of an artificial thermal conductivity can dramatically improve the ability of the SPH codes to exhibit this instability. There have since been a number of papers discussing this issue, but to our knowledge none of these improvements have yet been incorporated into an SPH model of a WD merger. Another reason to be skeptical is that the detonation itself was inserted manually in \cite{pakmor:2010} based on small-scale detonation simulations. A self-consistent evolution of a thermonuclear detonation, if it existed, would significantly build confidence in the progenitor model.
